Divergent morphology of elliptical bifaces from sites in the American Southeast articulates with raw material differences

### 2022-08-05 supp/index.Rmd
<edit readme>

--- 
title: "Caddo retouch results in divergent morphology for elliptical bifaces from sites in the American Southeast"
author:
    - Robert Z. Selden, Jr.^[Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E. Dockall^[Stantec, Inc.]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
description: "Pilot study of elliptical biface morphology."
---

@@

Elliptical bifaces from the Millsap Cache and Jowell Farm (41AN13) sites comprise two of the largest samples of this ill-understood lithic tool from the ancestral Caddo area. Similar to Gahagan bifaces, the bifaces from Jowell Farm are assumed to have been manufactured using Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade with central Texas groups. Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material and/or biogeographical range. The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (larger), and the other associated with resharpening and/or retouch (smaller). Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---knife shape at the smaller size class, and to test the hypothesis that greater morphological variation would be apparent in that size class due to wear associated with regular use, paired with variable idiosyncracies related to resharpening and/or retouch. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This finding is suggestive of a shared mental template, but a size differential that is potentially driven by raw material and/or biogeographical differences. Additionally, elliptical biface size associated with the large size class does not differ between sites/raw materials; however, the small size class does. This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening trajectory, where bifaces from each site/raw material enter their uselife at a similar shape and size, but diverge significantly in size---but not shape---throughout that uselife. In following with our expectations, greater shape diversity occurred in the small size class, where the bifaces from Jowell Farm were found to be more diverse than those from Millsap Cache. Discrete biogeographical ranges and raw material differences are posited as the drivers of these extant morphological differences found in elliptical bifaces from the ancestral Caddo area.

@@

The analysis code associated with this project can be accessed through this document, is available in the GitHub repository, and digitally curated on the Open Science Framework. The reproducible nature of this enterprise provides a means for others to critically assess and evaluate the various analytical components [@RN20915;@RN20916;@RN20917], which is a necessary requirement for the production of reliable knowledge.

>>>>

--- 
title: "Divergent morphology of elliptical bifaces from sites in the American Southeast articulates with raw material differences"
author:
    - Robert Z. Selden, Jr.^[Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E. Dockall^[Stantec, Inc.]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
description: "Pilot study of elliptical biface morphology."
---

@@

Elliptical bifaces from the Millsap Cache and Jowell Farm (41AN13) sites comprise two of the largest samples of this ill-understood lithic tool from the ancestral Caddo area. Similar to Gahagan bifaces, those bifaces from Jowell Farm are assumed to have been manufactured using Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade with central Texas groups. Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material. The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (larger), and the other associated with resharpening and/or retouch (smaller). Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---knife shape at the smaller size class, and to test the hypothesis that greater morphological variation would be apparent in the small size class due to the variable idiosyncracies related to resharpening and/or retouch. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This finding suggests both a shared mental template, and a size differential is potentially driven by raw material. Additionally, elliptical biface size in the large size class does not differ between raw materials; however, the small size class does. This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening trajectory, where bifaces from each site/raw material enter their uselife at a similar shape and size, diverging significantly in size---but not shape---throughout that uselife. In following with our expectations, greater shape diversity was found to occur in the small size class, where those bifaces from Jowell Farm were more diverse than those from the Millsap Cache. Discrete raw material differences are posited as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area.

@@

The analysis code associated with this project can be accessed through this document, is available in the GitHub repository [https://github.com/seldenlab/elliptical.bifaces](https://github.com/seldenlab/elliptical.bifaces), and digitally curated on the Open Science Framework [DOI 10.17605/OSF.IO/PH25W](https://osf.io/ph25w/). The reproducible nature of this enterprise provides a means for others to critically assess and evaluate the various analytical components [@RN20915;@RN20916;@RN20917], which is a necessary requirement for the production of reliable knowledge.
